Cognate beginnings to bilingual lexical acquisition

### 2023-09-19 manuscript/manuscript.tex
Update models with 'on_change'

source of this study. GGC was supported by a FPI 2020 research contract
(PRE2019-088165) in a project funded by the Spanish Ministry of Science,
Innovation, and Universities (PGC2018-101831-B-I00). DAV was supported
by the European Union's Horizon 2023 research and innovation program
under Marie Skłodowska--Curie Grant (765556) and a postdoctoral

@@

supported by Academia 2019 award from the Catalan Institution for

@@

equivalent to those shown by their monolingual peers. This is
remarkable, given the increased complexity of their linguistic input.
Recent studies suggested that bilingual vocabulary growth is boosted by
the number of cognates shared by the pair of languages being learned,
and that this facilitation effect is driven by a stronger parallel
activation of cognates during linguistic exposure, compared to
non-cognates. The mechanisms behind this facilitation are still unclear.
In this study, we capitalised on accumulator models of language
acquisition to propose an account of bilingual lexical acquisition in
which parallel activation increases the rate at which children
accumulate learning instances for words in both languages, even in fully
monolingual situations. Under this hypothesis, we predicted a stronger

@@

translation more often than high-exposure words do. We developed an

@@

bilingual toddlers aged 12 to 32 months. We then used Bayesian
explanatory item response theory to model the acquisition trajectories
of 302 Catalan/Spanish translation equivalents. We found an interaction
between exposure and cognateness, which pointed to cognateness
facilitating the aquisition of low-exposure words, but not of mean
exposure or high-exposure words. Overall, our findings suggest that
cognateness plays a key role in bilingual lexical acquisition, and
provides evidence for a frequency-mediated facilitation effect driven by
parallel activation.

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, interior hidden, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable, sharp corners]}{\end{tcolorbox}}\fi

@@

The foundations of word learning are in place early in age. At six
months, infants start directing their gaze to some objects when hearing
their labels (\protect\hyperlink{ref-bergelson2012months}{Bergelson \&
Swingley, 2012}, \protect\hyperlink{ref-bergelson2015early}{2015};
\protect\hyperlink{ref-tincoff1999beginnings}{Tincoff \& Jusczyk,
1999}), and shortly after caregivers start reporting some words as
acquired by their infant in vocabulary checklists (e.g.,
\protect\hyperlink{ref-fenson2007macarthurbates}{Fenson et al., 2007};
\protect\hyperlink{ref-samuelson2021precision}{Samuelson, 2021}). Most
research on early word acquisition relies extensively on data from
monolingual children, and is oblivious to the fact that a substantial
proportion of the world population acquires more than one language from
early ages (\protect\hyperlink{ref-grosjean2021extent}{Grosjean, 2021}).
Previous work on bilingual vocabulary acquisition pointed to bilingual
toddlers knowing, on average, less words in each of their languages than
their monolinguals peers, and to both groups knowing a similar number of
words---if not more words---when the bilinguals' two languages are
pooled together. Hoff et al.
(\protect\hyperlink{ref-hoff2012dual}{2012}) found that English-Spanish
bilingual toddlers in South Florida (United States) knew less words in
English than monolinguals did, but both groups knew a similar total
amount of words when both English and Spanish vocabularies were counted
together. Other studies have provided converging evidence that
bilinguals know a similar or even larger number of words than
monolinguals when the two languages are aggregated
(\protect\hyperlink{ref-gonzalez-barrero2020bilingual}{Gonzalez-Barrero
et al., 2020}; \protect\hyperlink{ref-oller2002language}{Oller \&
Eilers, 2002}; \protect\hyperlink{ref-patterson2004comparing}{Patterson,
2004}; \protect\hyperlink{ref-patterson2004bilingual}{Patterson \&
Pearson, 2004}; \protect\hyperlink{ref-pearson1993lexical}{Pearson et
al., 1993}; \protect\hyperlink{ref-pearson1994patterns}{Pearson \&
Fernández, 1994}; \protect\hyperlink{ref-petitto2001bilingual}{Petitto
et al., 2001}; \protect\hyperlink{ref-smithson2014bilingualism}{Smithson
et al., 2014}). A more detailed analysis of the words in bilinguals'
lexicon shows some interesting patterns.

One important observation of studies on bilinguals' early vocabulary
acquisition is that cognate words are easier to acquire than non-cognate
words. Cognate words are translation equivalents that are phonologically
similar (or share some type of form-similarity). For instance, the
Spanish translation equivalent of \emph{cat} is \emph{gato}, a cognate
word; the translation equivalent of \emph{dog} is \emph{perro}, a
non-cognate word. The differences in the percentages of cognate and
non-cognate words between two languages is related to historical
reasons: languages typologically close (like Dutch and English or
Italian and Spanish) share more cognates than languages typologically
distant (like English and Chinese, or Urdu and Spanish). The conclusion
that cognate words are easier to learn is based on two types of
evidence: studies investigating vocabulary sizes in children learning
language pairs with different percentages of cognates (that is,
differing in their typological distance) and studies comparing the
number of cognate and non-cognate words children know in a specific
language pair.

Floccia et al. (\protect\hyperlink{ref-floccia2018introduction}{2018})
published an impressive study comparing vocabularies of children
learning several language pairs differing in their percentage of
cognates. The authors collected vocabulary data on word comprehension
and production from 372 24-month-old bilingual toddlers living in the
United Kingdom who were learning English and an additional language. The
additional language was one of 13 typologically diverse languages:
Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi/Urdu,
Italian, Mandarin Chinese, Polish, Portuguese, Spanish, and Welsh. The
authors calculated the average lexical overlap between the words in each
of these additional languages and their translation equivalents in
English, which was taken as a proxy of the degree of cognateness between
each pair of languages. Floccia and co-workers reported an increase in
vocabulary size in the additional language (i.e., not English)
associated with an increase in the average phonological similarity
between the translation equivalents of each language pair. For example,
English-Dutch bilinguals (languages with a high phonological overlap),
were able to produce more Dutch words than English-Mandarin bilinguals
(languages with a low phonological overlap) were able to produce in
Mandarin. Blom et al.
(\protect\hyperlink{ref-blom2020crosslanguage}{2020}), Bosma et al.
(\protect\hyperlink{ref-bosma2019longitudinal}{2019}), and Gampe et al.
(\protect\hyperlink{ref-gampe2021does}{2021}) reported similar results,
providing converging evidence of a facilitatory effect of a lower
language distance (i.e., higher degre of cognateness) on vocabulary
size.

A second set of studies suggested that cognates are overrepresented in
bilinguals' early lexicon. Bosch \& Ramon-Casas
(\protect\hyperlink{ref-bosch2014first}{2014}) collected parental
reports of expressive vocabulary from 48 Catalan-Spanish bilinguals aged
18 months and found that cognates represented a larger proportion of
participant's vocabulary than non-cognates. Schelletter
(\protect\hyperlink{ref-schelletter2002effect}{2002}) provided
converging evidence from a longitudinal single-case study, in which an
English-German bilingual child produced cognates earlier than
non-cognates, on average. But the high proportion of cognates in the
vocabulary of the participants in these two studies may not necessarily
evidence of a facilitation effect of cognateness, but rather of simply
the high proportion of cognates present in the pair of languages being
learned. For instance, if two given languages share a high proportion of
cognates like \(\sim 70\)\%, the vocabulary contents of children
learning both languages should, in principle, approximate such
proportion of cognates, even in the absence of a cognateness
facilitation effect. More recently, Mitchell et al.
(\protect\hyperlink{ref-mitchell2022cognates}{2022}) addressed this
issue in a longitudinal study with a larger sample. The authors
collected expressive vocabulary data of 47 16- to 30-month-old
French-English bilinguals living in Canada, in both languages. They
created two lists of translation equivalents: one made of 131 cognates,
and one made of 406 non-cognates. The proportion of words that children
were reportedly able to produce was higher in the cognate lists than in
the non-cognate list across ages. Critically, this difference persisted
after both lists were matched in size, controlling their semantic
category (i.e., furniture, animals, food were similarity represented in
both lists) and age-of-acquisition norms (an index of word difficulty).
Taken together, the results of these two lines of research have been
considered as providing support to the hypothesis that phonological
similarity (as reflected in cognateness) plays a facilitation role in
bilingual word acquisition.

Parallel activation of bilinguals' lexicons has been proposed as the
underlying mechanism for such facilitatory effect (e.g.,
\protect\hyperlink{ref-floccia2018introduction}{Floccia et al., 2018};
\protect\hyperlink{ref-mitchell2022cognates}{Mitchell et al., 2022}).
The parallel activation hypothesis stems from the language non-selective
account of lexical access, which suggests that bilinguals activate both
languages simultaneously during language processing, even in fully
monolingual contexts. Evidence with adult bilinguals supporting the
language-non selective account of lexical access has been reported for
language comprehension and production, across the auditory and visual
(reading and signing) modalities
(\protect\hyperlink{ref-gimeno-martinez2021crosslinguistic}{Gimeno-Martínez
et al., 2021}; \protect\hyperlink{ref-vanhell2008sentence}{Hell \&
Groot, 2008}; \protect\hyperlink{ref-hoshino2008cognate}{Hoshino \&
Kroll, 2008}; \protect\hyperlink{ref-morford2011when}{Morford et al.,
2011}; \protect\hyperlink{ref-schwartz2007reading}{Schwartz et al.,
2007}; \protect\hyperlink{ref-spivey1999cross}{Spivey \& Marian, 1999};
see \protect\hyperlink{ref-kroll2017bilingual}{Kroll \& Ma, 2017} for
review). One of the clearest pieces of evidence of parallel activation
was provided by Costa et al.
(\protect\hyperlink{ref-costa2000cognate}{2000}). In this study, Spanish
monolinguals and Catalan-Spanish bilingual adults were asked to name
pictures of common objects in Spanish. In half of the trials, the object
labels were cognates in Spanish and Catalan (\emph{árbol}-\emph{arbre},
translations of \emph{tree}), whereas in the other half of the trial
labels were non-cognates (\emph{mesa}-\emph{taula}, translations of
\emph{table}), obviously, such distinction was only relevant for
bilinguals. Bilinguals named cognate pictures faster than non-cognate
pictures, even after adjusting for the lexical frequency of the items.
In contrast, Spanish monolinguals, who were unfamiliar with the Catalan
translations of the Spanish words they uttered, showed equivalent naming
times for the two types of stimuli. The authors interpreted the
difference between cognates and non-cognates in bilinguals as reflecting
the additional phonological activation that cognate words would receive
from their translation equivalents (due to language non-selective
activation of bilinguals' lexicons). These results showed that
bilinguals' Catalan phonology was activated during the production of
Spanish words, facilitating the naming of cognate pictures. Evidence of
parallel activation has been reported in bilingual toddlers and children
(\protect\hyperlink{ref-bosma2020cognate}{Bosma \& Nota, 2020};
\protect\hyperlink{ref-poarch2012crosslanguage}{Poarch \& Hell, 2012};
\protect\hyperlink{ref-poulin-dubois2013lexical}{Poulin-Dubois et al.,
2013}; \protect\hyperlink{ref-schroter2016orthographic}{Schröter \&
Schroeder, 2016}; \protect\hyperlink{ref-vonholzen2019impact}{Von Holzen
et al., 2019}; \protect\hyperlink{ref-vonholzen2012language}{Von Holzen
\& Mani, 2012}).

Although there is a consensus on the role of parallel activation in
bilinguals' lexical processing and acquisition, previous studies do not
address its influence on the learning trajectories of words. Results are
aggregated across words and provide no information about the specific
dynamics of how parallel activation influences word learning. This is
the goal of the present research.

We propose an account in which a learning instance for a word may also
represent a learning instance for its translation equivalent, to the
extent that such translation equivalent is co-activated\footnote{We use
  the term \emph{learning instance} in the fashion of accumulator models
  of language acquisition; as an exposure to a word-form that
  constitutes an opportunity for the child to accumulate information
  about the word. We do not make strong assumptions about whether a
  learning instance is a discrete or a continuous unit of accumulation
  of information. We rather consider that a learning instance of a word
  is so to the degree of the strength of activation of its phonological
  representation. This activation may result from the infant being
  exposed to the actual word-form, or the result of activation spreading
  through phonological or semantic links across lexical representation,
  as in the case of parallel activation.}. The strength of this
co-activation is proportional to the phonological similarity between the
two translation equivalents; given that cognates share higher
phonological similarity than non-cognates, the former should be
co-activated more strongly than the latter. This should lead to a faster
accumulation of learning instances for cognates, compared to
non-cognates. Parallel activation would allow bilingual children to
accumulate learning instances for words in both languages even during
fully monolingual situations, but the impact of this mechanism would be
asymmetric across languages: words from the lower-exposure language
would receive stronger activation from words in the higher-exposure
language than vice versa. Therefore, the acquisition of words from the
lower-exposure language would benefit more strongly from their cognate
status than words from the higher-exposure language. This asymmetric
cross-language activation would be consistent with previous reports of
larger priming effects from the dominant to the non-dominant language,

Consider the example of the Catalan-Spanish cognate translation
equivalent /\textipa{"gat}-\textipa{"ga.to}/ {[}\emph{cat}{]}, that are
phonologically very similar. When the child is exposed to
/\textipa{"gat}/, they will strongly co-activate /\textipa{"ga.to}/ in
parallel. Therefore, this exposure will count as a learning instance for
both co-activated forms. The case of the non-cognate translation
equivalent /\textipa{"gos-"pe.ro}/ {[}\emph{dog}{]} would be different.
Given the low phonological similarity between both word-forms, an
exposure to \textipa{"gos} will result in a weak activation of
/\textipa{"pe.ro}/ leading to such exposure counting as a learning
instance for /\textipa{"gos}/ (which the child was exposed to), but not
for /\textipa{"pe.ro}/. While /\textipa{"gat}-\textipa{"ga.to}/ will
benefit from parallel activation, /\textipa{"gos}-\textipa{"pe.ro}/ will
not. If the child receives linguistic input from one of the languages
more often than from the other, this effect might affect each form of
the cognate translation equivalent differently. For instance, if the
child receives a larger amount of Catalan input than Spanish input, they
will encounter the Catalan form /\textipa{"gat}/ more frequently than
the Spanish form /\textipa{"ga.to}/. Through parallel activation,
/\textipa{"gat}/ will activate /\textipa{"ga.to}/ more often than vice
versa. Ultimately, /\textipa{"ga.to}/ will benefit more strongly from
its cognate status than /\textipa{"gat}/, as it receives additional
learning instances from its translation equivalent more often than
/\textipa{"gat}/.

To test these predictions, we collected vocabulary data on production
and comprehension from a large sample of bilingual Catalan-Spanish
children using the Barcelona Vocabulary Questionnaire (BVQ)
(\protect\hyperlink{ref-garcia-castro2023bvq}{Garcia-Castro et al.,
2023}), a dual language online vocabulary checklist designed for the
present study. We adopted a Bayesian explanatory item response theory
approach (IRT, see
\protect\hyperlink{ref-kachergis2022standard}{Kachergis et al., 2022},
for a similar approach) to model the probability of acquisition of 604
Catalan and Spanish nouns included in the vocabulary checklist. Words
were considered as acquired if caregivers reported such word to be
understood (comprehension) or understood and said (production) by their
child. We estimated the impact of several predictors of interest on the
probability of acquisition, including participants' age and rate of
exposure to the word-form, and the cognate status of the word-form. As
described in the methods section, rate of exposure was a composite
measure taking into account participant' language exposure and word's
lexical frequency. We predicted an interaction between cognate status
and word exposure rate in which the probability of comprehension is
higher for low-exposure cognate words, but not for high-exposure cognate
words.

@@

To collect vocabulary data from participants, we created an \emph{ad
hoc} questionnaire: the Barcelona Vocabulary Questionnaire (BVQ)
(\protect\hyperlink{ref-garcia-castro2023bvq}{Garcia-Castro et al.,
2023}). This questionnaire was inspired by the MacArthur-Bates
Communicative Development Inventory
(\protect\hyperlink{ref-fenson2007macarthurbates}{Fenson et al., 2007})
and its adaptations to other languages, and was implemented on-line
using the formr platform (\protect\hyperlink{ref-arslan2020formr}{Arslan
et al., 2020}). This questionnaire is structured in three blocks: (1) a
language exposure questionnaire, (2) a demographic survey, and (3) two
vocabulary checklists. Vocabulary checklists followed a similar
structure as the Oxford Communicative Developmental Inventory (OCDI)
(\protect\hyperlink{ref-hamilton2000infant}{Hamilton et al., 2000}) and
consisted in two lists of words: one in Catalan and one in Spanish. Both
lists included items from a diverse sample of 26 semantic/functional
categories. The Catalan checklist contained 793 items and the Spanish
checklist contained 797. Items in one language were translation
equivalents of the items in the other language (e.g., the same
participant responded to both \emph{gos} and \emph{perro}, Catalan and
Spanish for \emph{dog}), roughly following a one-to-one
mapping\footnote{Although for some of the included words in Catalan did
  not have a clear translation or had more than one possible translation
  in Spanish, and vice versa, therefore the unequal number of words
  included in the Catalan and Spanish lists} (see Table~\ref{tbl-items}
for a summary of the questionnaire items).

@@

For each word included in the vocabulary checklists, we asked parents to
report whether their child was able to understand it, understand
\emph{and} say it, or did not understand or say it (checked out by
default). Given the large number of words in the vocabulary checklists,
we created four different subsets of the complete list of items. Each
subset contained a random but representative sub-sample of the items
from the complete list (see Table~\ref{tbl-items}). Semantic/functional
categories with less than 16 items---thus resulting in less than four
items after dividing it in four lists---were not divided in the short
version of the questionnaire: all of their items were included in the
four lists. Items that were part of the trial lists of some ongoing
experiments in the lab were also included in all versions. The resulting
reduced list contained between 343 and 349 Catalan words, and between
349 and 371 Spanish words. Participants were randomly allocated into one
of the four subsets.

@@

We implemented the model using \texttt{brms}
(\protect\hyperlink{ref-burkner2017brms}{Bürkner, 2017}), a R interface
to the Stan probabilistic language (2.32.1)
(\protect\hyperlink{ref-carpenter2017stan}{Carpenter et al., 2017}). We
ran four iteration chains using the by-default No U-Turn Sampler
algorithm with 1,000 iterations each and an additional 1,000 warm-up
iterations per chain. Model posterior draws and predictions were handled
using the \texttt{tidybayes}
(\protect\hyperlink{ref-kay2021tidybayes}{Kay, 2021}) R package.

>>>>

source of this study. This study was supported by the Spanish Ministry
for Science and Innovation and State Research Agency (Project PID2021-
123416NB-I00 financed by MCIN/ AEI/ 10.13039/501100011033 / FEDER, UE)
and the Economic and Social Research Council (ESRC) (ES/S010947/1, UK).
GGC was supported by a FPI research contract (PRE2019-088165). DAV was
supported by the European Union's Horizon 2023 research and innovation
program under Marie Skłodowska--Curie Grant (765556) and a postdoctoral

@@

supported by an ICREA Academia award from the Catalan Institution for

@@

equivalent to their monolingual peers'. This is remarkable, given the
complexity of their linguistic input. Recent studies suggest that
bilingual vocabulary growth is boosted by the number of cognates
(form-similar translation equivalents) shared by the pair of languages
being learned, and that this cognateness facilitation effect is driven
by a stronger parallel activation of cognates during linguistic
exposure, compared to non-cognates. The mechanisms behind this
facilitation are still unclear. In this study, we propose an account of
bilingual lexical acquisition in which parallel activation increases the
rate at which children accumulate learning instances for words in both
languages, even in fully monolingual situations. We predicted a stronger

@@

translation more often than high-exposure words. We developed an

@@

bilingual toddlers aged 12 to 32 months. We used Bayesian explanatory
item response theory to model the acquisition trajectories of 604
Catalan and Spanish words. We found an interaction between exposure and
cognateness, suggesting that cognateness facilitates the aquisition of
low-exposure words, but not of mean exposure or high-exposure words.
Overall, our findings suggest that cognateness plays a key role in
bilingual lexical acquisition, and provide evidence for a
frequency-mediated facilitation effect driven by parallel activation.

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, frame hidden, boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi

@@

The foundations of word learning are in place at an early age. At six
months, infants start directing their gaze to objects when hearing their
labels (\protect\hyperlink{ref-bergelson2012months}{Bergelson \&
Swingley, 2012}, \protect\hyperlink{ref-bergelson2015early}{2015};
\protect\hyperlink{ref-tincoff1999beginnings}{Tincoff \& Jusczyk,
1999}), and shortly after caregivers start reporting some words as
acquired by their infant in vocabulary checklists (e.g.,
\protect\hyperlink{ref-fenson2007macarthurbates}{Fenson et al., 2007};
\protect\hyperlink{ref-samuelson2021precision}{Samuelson, 2021}). Most
research on early word acquisition relies extensively on data from
monolingual children, and is oblivious to the fact that a substantial
proportion of the world population acquires more than one language from
early ages (\protect\hyperlink{ref-grosjean2021extent}{Grosjean, 2021}).
Previous work on bilingual vocabulary acquisition pointed to bilingual
toddlers knowing, on average, less words in each of their languages than
their monolinguals peers, and to both groups knowing a similar number of
words---if not more words---when the bilinguals' two languages are
pooled together. Hoff et al.
(\protect\hyperlink{ref-hoff2012dual}{2012}) found that English-Spanish
bilingual toddlers in South Florida (United States) knew less words in
English than monolinguals did, but both groups knew a similar total
amount of words when both English and Spanish vocabularies were counted
together. Other studies have provided converging evidence that
bilinguals know a similar or even larger number of words than
monolinguals when the two languages are aggregated
(\protect\hyperlink{ref-gonzalez-barrero2020bilingual}{Gonzalez-Barrero
et al., 2020}; \protect\hyperlink{ref-oller2002language}{Oller \&
Eilers, 2002}; \protect\hyperlink{ref-patterson2004comparing}{Patterson,
2004}; \protect\hyperlink{ref-patterson2004bilingual}{Patterson \&
Pearson, 2004}; \protect\hyperlink{ref-pearson1994patterns}{Pearson \&
Fernández, 1994};
\protect\hyperlink{ref-smithson2014bilingualism}{Smithson et al.,
2014}). A more detailed analysis of the words in bilinguals' lexicons
shows some interesting patterns.

One important observation of studies on bilinguals' early vocabulary
acquisition is that cognate words are easier to acquire than non-cognate
words. Cognate words are translation equivalents that are phonologically
similar (or share some type of form-similarity). For instance, the
Spanish translation equivalent of \emph{cat} is \emph{gato}, a cognate
word; the translation equivalent of \emph{dog} is \emph{perro}, a
non-cognate word. For historical reasons, some pairs of languages share
more cognates than others: languages typologically close (like Dutch and
English or Italian and Spanish) share more cognates than languages
typologically distant (like English and Chinese, or Urdu and Spanish).
The conclusion that cognate words are easier to learn is based on two
types of evidence: studies investigating vocabulary sizes in children
learning language pairs with different percentages of cognates (that is,
differing in their typological distance) and studies comparing the
number of cognate and non-cognate words children know in a specific
language pair.

Floccia et al. (\protect\hyperlink{ref-floccia2018introduction}{2018})
published an impressive study comparing vocabularies of children
learning several language pairs differing in their percentage of
cognates. The authors collected vocabulary data on word comprehension
and production from 372 24-month-old bilingual toddlers living in the
United Kingdom who were learning English and an additional language. The
additional language was one of 13 typologically diverse languages:
Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi or Urdu,
Italian, Mandarin Chinese, Polish, Portuguese, Spanish, and Welsh. The
authors calculated the average lexical overlap between the words in each
of these additional languages and their translation equivalents in
English. Lexical overlap was calculated in terms of phonological
similarity (described below) and it was taken as a proxy of the degree
of cognateness between each pair of languages. Floccia and co-workers
reported an increase in vocabulary size in the additional language
(i.e., not English) associated with an increase in the average
phonological similarity between the translation equivalents of each
language pair. For example, English-Dutch bilinguals (languages with a
high phonological overlap), were able to produce more Dutch words than
English-Mandarin bilinguals (languages with a low phonological overlap)
were able to produce in Mandarin. Blom et al.
(\protect\hyperlink{ref-blom2020crosslanguage}{2020}), Bosma et al.
(\protect\hyperlink{ref-bosma2019longitudinal}{2019}), and Gampe et al.
(\protect\hyperlink{ref-gampe2021does}{2021}) reported similar results,
providing converging evidence of a facilitatory effect of a lower
language distance (i.e., higher degre of cognateness) on vocabulary
size.

A second set of studies suggested that cognates are overrepresented in
bilinguals' early lexicon. Bosch \& Ramon-Casas
(\protect\hyperlink{ref-bosch2014first}{2014}) collected parental
reports of expressive vocabulary from 48 Catalan-Spanish bilinguals aged
18 months and found that cognates represented a larger proportion of
vocabulary than non-cognates. Schelletter
(\protect\hyperlink{ref-schelletter2002effect}{2002}) provided
converging evidence from a longitudinal single-case study, in which an
English-German bilingual child produced cognates earlier than
non-cognates, on average. But the high proportion of cognates in the
vocabulary of the participants in these two studies may not necessarily
evidence of a facilitation effect of cognateness, but rather of simply
the high proportion of cognates present in the pair of languages being
learned. For instance, if two given languages share a high proportion of
cognates like 70\%, the vocabulary contents of children learning both
languages should, in principle, approximate such proportion of cognates,
even in the absence of a cognateness facilitation effect. More recently,
Mitchell et al. (\protect\hyperlink{ref-mitchell2022cognates}{2022})
addressed this issue in a longitudinal study. The authors collected
expressive vocabulary data of 47 16- to 30-month-old French-English
bilinguals living in Canada, in both languages. They created two lists
of translation equivalents: one made of 131 cognates, and one made of
406 non-cognates. The proportion of words that children were reportedly
able to produce was higher in the cognate lists than in the non-cognate
list across ages. Critically, this difference persisted after both lists
were matched in size, controlling their semantic category (i.e.,
furniture, animals, food were similarity represented in both lists) and
age-of-acquisition norms (an index of word difficulty). Taken together,
the results of these two lines of research have been considered as
supporting to the hypothesis that phonological similarity (as reflected
in cognateness) plays a facilitation role in bilingual word acquisition.

Parallel activation of bilinguals' lexicons has been proposed as the
underlying mechanism for such facilitatory effect (e.g.,
\protect\hyperlink{ref-floccia2018introduction}{Floccia et al., 2018};
\protect\hyperlink{ref-mitchell2022cognates}{Mitchell et al., 2022}).
The parallel activation hypothesis stems from the language non-selective
account of lexical access, which suggests that bilinguals activate both
languages simultaneously during language processing, even in fully
monolingual contexts. Evidence with adult bilinguals supporting the
language-non selective account of lexical access has been reported for
language comprehension and production, across the auditory and visual
(reading and signing) modalities
(\protect\hyperlink{ref-gimeno-martinez2021crosslinguistic}{Gimeno-Martínez
et al., 2021}; \protect\hyperlink{ref-hoshino2008cognate}{Hoshino \&
Kroll, 2008}; \protect\hyperlink{ref-morford2011when}{Morford et al.,
2011}; \protect\hyperlink{ref-shook2012bimodal}{Shook \& Marian, 2012};
\protect\hyperlink{ref-spivey1999cross}{Spivey \& Marian, 1999}; see
\protect\hyperlink{ref-kroll2017bilingual}{Kroll \& Ma, 2017} for
review). One of the clearest pieces of evidence of parallel activation
was provided by Costa et al.
(\protect\hyperlink{ref-costa2000cognate}{2000}). In this study, Spanish
monolinguals and Catalan-Spanish bilingual adults were asked to name
pictures of common objects in Spanish. In half of the trials, the object
labels were cognates in Spanish and Catalan (\emph{árbol}-\emph{arbre},
translations of \emph{tree}), whereas in the other half of the trial
labels were non-cognates (\emph{mesa}-\emph{taula}, translations of
\emph{table}). Obviously, such distinction was only relevant for
bilinguals. Bilinguals named cognate pictures faster than non-cognate
pictures, even after adjusting for the lexical frequency of the items.
In contrast, Spanish monolinguals, who were unfamiliar with the Catalan
translations of the Spanish words they uttered, showed equivalent naming
times for the two types of stimuli. The authors interpreted the
difference between cognates and non-cognates in bilinguals as reflecting
the additional phonological activation that cognate words would receive
from their translation equivalents (due to language non-selective
activation of bilinguals' lexicons). These results showed that
bilinguals' Catalan phonology was activated during the production of
Spanish words, facilitating the naming of cognate pictures. More
recntly, evidence of parallel activation has been reported in bilingual
toddlers and children too
(\protect\hyperlink{ref-bosma2020cognate}{Bosma \& Nota, 2020};
\protect\hyperlink{ref-floccia2020translation}{Floccia et al., 2020};
\protect\hyperlink{ref-poarch2012crosslanguage}{Poarch \& Hell, 2012};
\protect\hyperlink{ref-poulin-dubois2013lexical}{Poulin-Dubois et al.,
2013}; \protect\hyperlink{ref-vonholzen2019impact}{Von Holzen et al.,
2019}; \protect\hyperlink{ref-vonholzen2012language}{Von Holzen \& Mani,
2012}). Although there is a consensus on the role of parallel activation
in bilinguals' lexical processing and acquisition, previous studies do
not address its influence on the learning trajectories of words. Results
are aggregated across words and provide no information about the
specific dynamics of how parallel activation influences word learning.
This is the goal of the present research.

We propose an account in which a learning instance for a word may also
represent a learning instance for its translation equivalent, to the
extent that such translation equivalent is co-activated. We use the term
\emph{learning instance} in the fashion of accumulator models of
language acquisition; as an exposure to a word-form that constitutes an
opportunity for the child to accumulate information about the word. We
do not assume if a learning instance is a discrete or a continuous unit
of accumulation of information. We consider that a learning instance of
a word is an exposure to its (phonological) form if the resulting
strength of activation of its representation in the child's lexicon
reaches some thoretical threshold that leads to word-form recognition.
This activation may result from the infant being exposed to the actual
word-form, or the result of activation spreading through phonological or
semantic links across lexical representation, as in the case of parallel
activation. The strength of this co-activation is proportional to the
phonological similarity between the two translation equivalents; given
that cognates share higher phonological similarity than non-cognates,
the former should be co-activated more strongly than the latter. This
should lead to a faster accumulation of learning instances for cognates,
compared to non-cognates. Parallel activation would allow bilingual
children to accumulate learning instances for words in both languages
even during fully monolingual situations, but the impact of this
mechanism would be asymmetric across languages: words from the
lower-exposure language would receive stronger activation from words in
the higher-exposure language than vice versa. Therefore, the acquisition
of words from the lower-exposure language would benefit more strongly
from their cognate status than words from the higher-exposure language.
This asymmetric cross-language activation would be consistent with
previous reports of larger priming effects from the dominant to the
non-dominant language.

Consider the example of the Catalan-Spanish cognate translation
equivalent /\textipa{"gat}/--/\textipa{"ga.to}/ {[}\emph{cat}{]}, which
are phonologically very similar. When the child listens to
/\textipa{"gat}/, they will strongly co-activate /\textipa{"ga.to}/ in
parallel. If the child has already formed a form-meaning association for
both word-forms, parallel activation may result from the activation of
their common concept or from activation spreading throught phonological
similarity. We assume semantic co-activation to be constant across
cognate and non-cognate translation equivalents, and focus on
phonological co-activation as an additional source of activation that
affects cognates more strongly than non-cognates. Therefore, this
exposure will count as a learning instance for both co-activated forms.
The case of the non-cognate translation equivalent
/\textipa{"gos-"pe.ro}/ {[}\emph{dog}{]} would be different. Given the
low phonological similarity between both word-forms, an exposure to
\textipa{"gos} will result in a weak activation of /\textipa{"pe.ro}/
leading to such exposure counting as a learning instance for
/\textipa{"gos}/ (which the child was exposed to), but not for
/\textipa{"pe.ro}/. While /\textipa{"gat}/--/\textipa{"ga.to}/ will
benefit from phonological co-activation,
/\textipa{"gos}-\textipa{"pe.ro}/ will not. If the child receives
linguistic input from one of the languages more often than from the
other, this effect might affect each form of the cognate translation
equivalent differently. For instance, if the child receives a larger
amount of Catalan input than Spanish input, they will encounter the
Catalan form /\textipa{"gat}/ more frequently than the Spanish form
/\textipa{"ga.to}/. Through parallel activation, /\textipa{"gat}/ will
activate /\textipa{"ga.to}/ more often than vice versa. Ultimately,
/\textipa{"ga.to}/ will benefit more strongly from its cognate status
than /\textipa{"gat}/, as it receives additional learning instances from
its translation equivalent more often than /\textipa{"gat}/.

To test these predictions, we collected vocabulary data on production
and comprehension from a large sample of bilingual Catalan-Spanish
children. We adopted a Bayesian explanatory item response theory
approach to model the probability of acquisition of 604 Catalan and
Spanish nouns included in the vocabulary checklist. Words were
considered as acquired if caregivers reported such word to be understood
(comprehension) or understood and said (production) by their child. We
estimated the impact of several predictors of interest on the
probability of acquisition, including participants' age and rate of
exposure to the word-form, and the cognate status of the word-form. As
described in the methods section, rate of exposure was a composite
measure taking into account participant' language exposure and word's
lexical frequency. We predicted an interaction between cognate status
and word-form exposure rate in which the probability of comprehension is
higher for low-exposure cognate words, but not for high-exposure cognate
words.

@@

To collect vocabulary data from participants, we created an \emph{ad
hoc} questionnaire: the Barcelona Vocabulary Questionnaire (BVQ)
(\protect\hyperlink{ref-garcia-castro2023bvq}{Garcia-Castro et al.,
2023}). This questionnaire was inspired by the MacArthur-Bates
Communicative Development Inventory
(\protect\hyperlink{ref-fenson2007macarthurbates}{Fenson et al., 2007})
and its adaptations to other languages, and was implemented on-line
using the \texttt{formr} platform
(\protect\hyperlink{ref-arslan2020formr}{Arslan et al., 2020}). This
questionnaire is structured in three blocks: (1) a language exposure
questionnaire, (2) a demographic survey, and (3) two vocabulary
checklists. Vocabulary checklists followed a similar structure as the
Oxford Communicative Developmental Inventory
(\protect\hyperlink{ref-hamilton2000infant}{Hamilton et al., 2000}) and
consisted in two lists of words: one in Catalan and one in Spanish. Both
lists included items from a diverse sample of 26 semantic or functional
categories. The Catalan checklist contained 793 items and the Spanish
checklist contained 797. Items in one language were translation
equivalents of the items in the other language (e.g., the same
participant responded to both \emph{gos} and \emph{perro}, Catalan and
Spanish for \emph{dog}), roughly following a one-to-one mapping. Some of
the words in Catalan did not have a clear translation or had more than
one possible translation in Spanish, and vice versa, therefore the
unequal number of words included in the two lists (see
Table~\ref{tbl-items} for a summary of the questionnaire items).

@@

For each word included in the vocabulary checklists, we asked parents to
report whether their child was able to understand it, understand
\emph{and} say it, or did not understand or say it (checked out by
default). Given the large number of words in the vocabulary checklists,
we created four different subsets of the complete list of items: A, B,
C, and D. Each subset contained a random but representative sub-sample
of the items from the complete list (see Table~\ref{tbl-items}).
Semantic or functional categories with less than 16 items---thus
resulting in less than four items after dividing it in four lists---were
not divided: all of their items were included in the four lists. Items
that were part of the trial lists of some ongoing experiments in the lab
were also included in all versions. The resulting reduced list contained
between 343 and 349 Catalan words, and between 349 and 371 Spanish
words. Participants were randomly allocated into one of the four
subsets.

@@
